---
title: _signatureSearchData_ - Reference Data for Large-Scale Gene Expression Signature Searching
author: "Authors: Yuzhu Duan and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 4
    fig_caption: yes
    code_folding: show
    number_sections: true
  pdf_document:
    toc: true
fontsize: 15pt
bibliography: bibtex.bib
vignette: >
  %\VignetteIndexEntry{signatureSearchData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
    library(signatureSearchData)
})
# knitr::opts_knit$set(root.dir = "~/insync/project/longevityTools_eDRUG/")
```

# Introduction

The `signatureSearchData` package provides access to the reference data used by
the associated `signatureSearch` software package. The latter allows to search 
with a query gene expression signature (GES) a database of treatment GESs to 
identify cellular states with similar expression responses (connections). This 
way one can identify drugs or gene knockouts that induce expression phenotypes 
similar to a sample of interest. The resulting associations may lead to novel 
functional insights how perturbagens of interest interact with biological systems. 

Currently, `signatureSearchData` includes gene expression data from the CMap
(Connectivity Map) and LINCS (Library of Network-Based Cellular Signatures)
projects that are largely based on drug and genetic perturbation experiments
performed on variable numbers of human cell lines [@Lamb2006-du;
@Subramanian2017-fu]. In `signatureSearchData` these data sets have been
preprocessed for the gene expression signature search (GESS) algorithms
implemented in `signatureSearch`. The preprocessed data types include but are
not limited to normalized gene expression values (_e.g._ intensity values), log
fold changes (LFC) and Z-scores, p-values or FDRs of differentially expressed 
genes (DEGs), rankings based on selected preprocessing routines or sets of top 
up/down-regulated DEGs. 

The CMap data were downloaded from the [CMap project
site](https://portals.broadinstitute.org/cmap/) (Version build02). The latter is
a collection of over 7,000 gene expression profiles (signatures) obtained
from perturbation experiments with 1,309 drug-like small molecules on five
human cancer cell lines. The Affymetrix Gene Chip technology was used to
generate the CMAP2 data set. 

In 2017, the LINCS Consortium generated a similar but much larger data set where
the total number of gene expression signatures was scaled up to over one
million. This was achieved by switching to a much more cost effective gene
expression profiling technology called L1000 assay [@Peck2006-rf;
@Edgar2002-di]. The current set of perturbations covered by the LINCS data set
includes 19,811 drug-like small molecules applied at variable concentrations
and treatment times to ~70 human non-cancer (normal) and cancer cell lines.
Additionally, it includes several thousand genetic perturbagens composed of
gene knockdown and over-expression experiments. 

The data structures and search algorithms used by `signatureSearch` and
`signatureSearchData` are designed to work with most genome-wide expression
data including hybridization-based methods, such as Affymetrix or L1000, as
well as sequencing-based methods, such as RNA-Seq. Currently,
`signatureSearchData` does not include RNA-Seq data mainly due to the lack
of large-scale perturbation studies (_e.g._ drug-based) available in the public
domain that are based on RNA-Seq. This situation may change in the near future
once the technology has become more affordable. 


# Install and Load Package

`signatureSearch` is a R/Bioconductor package and can be installed using 
`BiocManager::install()`.

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("signatureSearchData", version = "3.9")
```

After the package is installed, it can be loaded in an R session as follows.
```{r load, eval=FALSE}
library(signatureSearchData)
```

# LINCS Signature Database

The L1000 assay, used for generating the LINCS data, measures the expression 
of 978 landmark genes and 80 control genes by loading amplified mRNA populations 
onto beads and then detecting their abundance with a fluorescent-based method [@Peck2006-rf]. 
The expression of 11,350 additional genes is imputed from the landmark genes by using 
as training data a large collection of Affymetrix gene chips [@Edgar2002-di].

The LINCS data have been pre-processed by the Broad Institute to 5 different levels
and are available for download from GEO. Level 1 data are the raw mean
fluorescent intensity values that come directly from the Luminex scanner. Level
2 data are the expression intensities of the 978 landmark genes. They have been
normalized and used to impute the expression of the additional 11,350 genes,
forming Level 3 data. A robust z-scoring procedure was used to generate
differential expression values from the normalized profiles (Level 4).
Finally, a moderated z-scoring procedure was applied to the replicated samples
of each experiment (mostly 3 replicates) to compute a weighted average
signature (Level 5). For a more detailed description of the preprocessing
methods used by the LINCS project, readers want to refer to the [LINCS user
guide](https://docs.google.com/document/d/1q2gciWRhVCAAnlvF2iRLuJ7whrGP6QjpsCMq1yWz7dU/edit#).

Disregarding replicates, the LINCS data set contains 473,647 signatures with
unique cell type and treatment combinations. This includes 19,811 drug-like
small molecules tested on different cell lines at multiple concentrations and
treatment times. In addition to compounds, several thousand genetic
perturbations (gene knock-downs and over expressions) have been tested.
Currently, the data described in this vignette are restricted to signatures of
small molecule treatments across different cells lines. However, users have the
option to assemble any custom collection of the LINCS data. For consistency,
only signatures at one specific concentration (10$\mu$M) and one time point
(24h) have been selected for each small molecule in the default collection.
These choices are similar to the conditions used in primary high-throughput
compound screens of cell lines.  Since the selected compound concentrations and
treatment duration have not been tested by LINCS across all cell types yet, a
subset of compounds had to be selected that best met the chosen treatment
requirements. This left us with 8,104 compounds that were uniformly tested at
the chosen concentration and treatment time, but across variable numbers of
cell lines. The total number of expression signatures meeting this requirement
is 45,956, while the total number of cell lines included in this data set is
30. 

## Explore Data Sets

A summary of the data sets provided by the `signatureSearchData` package can be
obtained with the `query` function of the `AnnotationHub` package. The information
is stored in an object of class `AnnotationHub`, here assigned to `ssd`.

```{r ah_explore_ssd1, eval=TRUE}
library(AnnotationHub)
ah <- AnnotationHub()
ssd <- query(ah, c("signatureSearchData"))
ssd
```

More detailed information about each data set can be returned as a `list`, here 
subsetted to first entry with `[1]`.

```{r ah_explore_ssd2, eval=TRUE}
as.list(ssd)[1]
```


## Z-scores from `AnnotationHub`

The LINCS sub-dataset, filtered and assembled according to the above criteria,
can be downloaded from Bioconductor's `AnnotationHub` as HDF5 file. In the
example below, the path to this file is assigned to a character vector called
`lincs_db_path`. A summary of the content of the HDF5 file can be returned with
the `h5ls` function. Note, due to the large size of the LINCS data set, its download 
takes too much time to evaluate the following code section during the build time of 
this vignette. 

```{r lincs, eval=FALSE}
library(AnnotationHub); library(rhdf5)
ah <- AnnotationHub()
query(ah, c("signatureSearchData", "lincs"))
lincs_db_path <- ah[['AH69092']]
h5ls(lincs_db_path) 
```

In this case the loaded data instance includes moderated Z-scores from DE
analyses of 12,328 genes for 8,140 compound treatments across a total of 30
cell lines corresponding to 45,956 expression signatures. This data set can be
used by all set-based GESS methods provided by the `signatureSearch` package.

## Z-scores from GEO 

The following explains how to generate the above LINCS data object from
scratch. This also illustrates how to filter the LINCS level 5 data in other
ways.

### Download Level 5 Data 

Download and unzip the following files from GEO entry [GSE92742](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742):

+ <tt>GSE92742_Broad_LINCS_Level5_COMPZ.MODZ_n473647x12328.gctx.gz</tt>
+ <tt>GSE92742_Broad_LINCS_gene_info.txt.gz</tt>
+ <tt>GSE92742_Broad_LINCS_sig_info.txt.gz</tt>

The following code examples assume that the downloaded datasets are stored in a
sub-directory called `data`. All paths in this vignette are given relative to
the present working directory of a user's R session.

### Filter Signatures 

The following selects LINCS Level 5 signatures of compound treatments at a 
concentration of 10$\mu$M and a treatment time of 24 hours. Note, the import command
below may issue a warning message that can be ignored. 

```{r filter_meta42, eval=FALSE}
meta42 <- readr::read_tsv("./data/GSE92742_Broad_LINCS_sig_info.txt") 
dose <- meta42$pert_idose[7]
## filter rows by 'pert_type' as compound, 10uM concentration, and 24h treatment time
meta42_filter <- sig_filter(meta42, pert_type="trt_cp", dose=dose, time="24 h") # 45956 X 14
```

### Z-score Data in HDF5

Next, the large Z-score matrix of expression signatures is imported step-wise
in subsets of manageable sizes and then appended to an HDF5 file (here
`lincs.h5`).  In this vignette, the latter is referred to as the LINCS
Z-score database. Since the size of the full matrix is several GBs in size, it would
consume too much memory to be read it into R at once. Reading the matrix in
smaller batches and appending them to an HDF5 file is much more memory
efficient. Subsequently, the `readHDF5chunk` function, defined by the
`signatureSearchData` package, imports user-definable subsets of the data from
the HDF5 file into a `SummarizedExperiment` object, here assigned to `se`.

```{r extract_modz, eval=FALSE}
gctx <- "./data/GSE92742_Broad_LINCS_Level5_COMPZ.MODZ_n473647x12328.gctx"
gctx2h5(gctx, cid=meta42_filter$sig_id, new_cid=meta42_filter$pert_cell_factor,
        h5file="./data/lincs.h5", chunksize=5000, overwrite=TRUE)
se <- readHDF5chunk(h5file="./data/lincs.h5", colindex=1:5000)
```

## Intensities from `AnnotationHub`

The LINCS Level 3 data can be downloaded from
[GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742) the same way
as described above for the Level 5 data. The Level 3 data contain normalized
gene expression values across all treatments and cell lines used by LINCS.  The
Level 3 signatures were filtered using the same dosage and duration criteria as
the Level 5 data. The biological replicate information included in the Level 3
data were collapsed to mean values.  Subsequently, the resulting matrix of mean
expression values was written to an HDF5 file. The latter is referred to as
`lincs_expr` database containing 38,824 signatures for a total of 5,925 small
molecule treatments and 30 cell lines. Although the LINCS Level 3 and 5 data
are filtered here the same way, the number of small molecules represented in
the Level 3 data (5,925) is smaller than in the Level 5 data (8,140). The reason for
this inconsistency is most likely that the Level 3 dataset, downloadable from GEO,
is incomplete.

The filtered and processed LINCS Level3 data (`lincs_expr`) can be loaded from 
Bioconductor's `AnnotationHub` interface as follows.

```{r lincs_expr, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("signatureSearchData", "lincs_expr"))
lincs_expr_path <- ah[['AH69093']]
```

In this case the loaded `lincs_expr` instance includes mean expression values of 
12,328 genes for 5,925 compound treatments across a total of 30 cell lines. 
This data set can be used by all correlation-based GESS methods provided by the 
`signatureSearch` package.

## Intensities from GEO

The following steps explain how to generate the above data set from scratch. 
This also illustrates how to filter the LINCS Level 3 data in other ways.

### Download Level 3 Data 

Download and unzip the following files from GEO entry [GSE92742](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742):

+ <tt>GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx.gz</tt>
+ <tt>GSE92742_Broad_LINCS_gene_info.txt.gz</tt>
+ <tt>GSE92742_Broad_LINCS_inst_info.txt.gz</tt>

As above, the following code examples assume that the downloaded datasets are
stored in a sub-directory called `data`. All paths in this vignette are given
relative to the present working directory of a user's R session.

### Filter Signatures 

The following selects LINCS Level 3 signatures of compound treatments at a 
concentration of 10$\mu$M and a treatment time of 24 hours. 

```{r filter_expr, eval=FALSE}
inst42 <- readr::read_tsv("./data/GSE92742_Broad_LINCS_inst_info.txt") 
inst42_filter <- inst_filter(inst42, pert_type="trt_cp", dose=10, dose_unit="um", 
                             time=24, time_unit="h") # 169,795 X 13
```

### Mean Intensities in HDF5

Next, mean expression values are calculated among biological replicates and then appended
in batches to the corresponding HDF5 file.

```{r extract_expr, eval=FALSE}
# It takes some time
meanExpr2h5(gctx="./data/GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx",
            inst=inst42_filter, h5file="./data/lincs_expr.h5") # 12328 X 38824
```

# CMap2 Signature Database

CMap2 (Version build02) contains GESs for 1,309 drugs and eight cell lines that
were generated with Affymetrix Gene Chips as expression platform. In some cases
this includes drug treatments at different concentrations and time points.  For
consistency, the CMap2 data was reduced to drug treatments with concentrations
and time points that are comparable to those used for the above LINCS data.
CMap2 data can be downloaded from GEO or its project site either in raw format or
as rank transformed matrix.  The ranks are based on DEG analyses of drug
treatments (drug vs. no-drug) where the resulting Z-scores were used to
generate the rank matrix. The latter was used here and is referred to as
`rankMatrix`. The Affymetrix probe set identifiers stored in the row name slot 
of this matrix were translated into gene identifies.  To obtain a matrix with unique gene
identifiers, the ranks for genes represented by more than one probe set were
averaged and then re-ranked accordingly. This final gene level rank matrix, referred to as `cmap_rank`,
contains rank profiles for 12,403 genes from 1,309 compound treatments in up to 5
cells corresponding to a total of 3,587 treatment signatures. This matrix can
be used for all GESS methods in the `signatureSearch` package that are
compatible with rank data, such as the `gess_cmap` method.

## Rank Matrix from `AnnotationHub`

The `cmap_rank` data can be downloaded from Bioconductor's `AnnotationHub` as HDF5
file. Since CMap2 is much smaller than LINCS, it can be imported in its
entirety into a `SummarizedExperiment` object (here assigned to `se`) without
excessive memory requirements. 

```{r cmap_rank, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("signatureSearchData", "cmap_rank"))
cmap_rank_path <- ah[["AH69763"]]
se <- readHDF5chunk(h5file=cmap_rank_path, colindex=1:3587)
```

## Rank Matrix from Sources

The following steps explain how to generate the above CMap2 rank data set from scratch. 

### Download Rank Data 

The `rankMatrix` can be downloaded from the CMap project site [here](https://portals.broadinstitute.org/cmap).
The specific file to download from this site is [rankMatrix.txt.zip](ftp://ftp.broad.mit.edu/pub/cmap/rankMatrix.txt.zip).
As before, it should be saved and unzipped in the `data` directory of a user's R session.

### Filter instances

The following selects from `rankMatrix` for each compound the chosen treatment
concentration and time point. This is achieved with help of the experiment
annotation file `cmap_instances_02.txt`, also available from the CMap project
site. Since this file is relatively small it has been included in the
`signatureSearchData` package from where it can be loaded into R as shown
below.

```{r filter_rankm, eval=FALSE}
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
inst_id <- cmap_inst$instance_id[!duplicated(paste(cmap_inst$cmap_name, cmap_inst$cell2, sep="_"))]
rankM <- read.delim("./data/rankMatrix.txt", check.names=FALSE, row.names=1) # 22283 X 6100
rankM_sub <- rankM[, as.character(inst_id)]
colnames(rankM_sub) <- unique(paste(cmap_inst$cmap_name, cmap_inst$cell2, "trt_cp", sep="__"))
```

### Annotated Gene Level Data

#### Annotation Information

The following generates annotation information for Affymetirx probe set
identifiers. Note, the three different Affymetrix chip types (HG-U133A,
HT_HG-U133A, U133AAofAv2) used by CMap2 share nearly all probe set identifiers,
meaning it is possible to use the same annotation package (here `hgu133a.db`)
for all three.

```{r affyid_annot, eval=FALSE, message=FALSE}
library(hgu133a.db)
myAnnot <- data.frame(ACCNUM=sapply(contents(hgu133aACCNUM), paste, collapse=", "), 
                      SYMBOL=sapply(contents(hgu133aSYMBOL), paste, collapse=", "), 
                      UNIGENE=sapply(contents(hgu133aUNIGENE), paste, collapse=", "), 
                      ENTREZID=sapply(contents(hgu133aENTREZID), paste, collapse=", "), 
                      ENSEMBL=sapply(contents(hgu133aENSEMBL), paste, collapse=", "), 
                      DESC=sapply(contents(hgu133aGENENAME), paste, collapse=", "))
saveRDS(myAnnot, "./data/myAnnot.rds")
```

#### Gene Level Data

The `probe2gene` function transforms probe set to gene level data. If genes are
represented by several probe sets then their mean intensities are used.

```{r mr_prob, eval=FALSE}
rankM_sub_gene <- probe2gene(rankM_sub, myAnnot) 
```

### Store Rank Matrix in HDF5 file

The sub-setted `rankMatrix` is written to an HDF5 file, referred to as
`cmap_rank` database.

```{r cmap_rank2h5, eval=FALSE}
matrix2h5(rankM_sub_gene, "./data/cmap_rank.h5", overwrite=TRUE) # 12403 X 3587
rhdf5::h5ls("./data/cmap_rank.h5")
## Read in cmap_rank.h5 as SummarizedExperiment object by chunks
cmap_rank_se <- readHDF5chunk("./data/cmap_rank.h5", colindex=1:5)
```

<font color="blue">[ThG-Comment: unclear where cmap_rank.h5 is generated. Without it the code is not 
reproducible.]</font>

## Intensities from `AnnotationHub`

To search CMap2 with `signatureSearch's` correlation based GESS methods
(`gess_cor`), normalized gene expression values (here intensities) are required
where the biological replicate information has been collapsed to mean values.
For this, the `cmap_expr` database has been created from CEL files, which are
the raw data of the Affymetrix technology. To obtain normalized expression
data, the CEL files were downloaded from the CMap project site, and then
processed with the MAS5 algorithm. Gene level expression data was generated the
same way as described above. Next, the gene expression values for different
concentrations and treatment times of each compound and cell were averaged.
Subsequently, the expression matrix was saved to an HDF5 file, 
referred to as the `cmap_expr` database. It represents mean expression values
of 12,403 genes for 1,309 compound treatments in up to 5 cells (3,587 signatures in total).

The `cmap_expr` database can be downloaded as HDF5 file from Bioconductor's 
`AnnotationHub` as follows.

```{r cmap_expr, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("signatureSearchData", "cmap_expr"))
cmap_expr_path <- ah[["AH69091"]]
```

This data set can be used by all correlation-based GESS methods provided by the
`signatureSearch` package.

## Intensities from Sources
The following steps explain how to generate the above `cmap_expr` database from scratch. 

### Directory Structure

The large number of files processed in the next steps are organized in two
sub-directories of a user's R session. Input files will be stored in a `data`
directory, while all output files will be written to a `results` directory. 

```{r work_envir, eval=FALSE}
dir.create("data"); dir.create("data/CEL"); dir.create("results") 
```

### Download of CEL Files

The `getCmapCEL` function will download the 7,056 CEL files from the [CMap
project site](http://www.broadinstitute.org/cmap), and save each of them to a
subdirectory named `CEL` under `data`. Since this download step will take some time,
the argument `rerun` has been assigned `FALSE` in the below function call to
avoid running it accidentally. To execute the download, the argument `rerun`
needs to be assigned `TRUE`. If the raw data are not needed, users can skip
this time consuming step and work with the preprocessed `cmap_expr` database
downloaded from the `AnnotationHub` instead.

```{r download_cmap, eval=FALSE}
getCmapCEL(rerun=FALSE) 
```

### Determine Chip Type 
The CMAP data set is based on three different Affymetrix chip types (HG-U133A,
HT_HG-U133A and U133AAofAv2). The following extracts the chip type information
from the downloaded CEL files and stores the information in an `rds` file with the path 
`./data/chiptype.rds`.

```{r get_cel_type, eval=FALSE}
library(affxparser)
celfiles <- list.files("./data/CEL", pattern=".CEL$")
chiptype <- sapply(celfiles, function(x) affxparser::readCelHeader(paste0("data/CEL/", x))$chiptype)
saveRDS(chiptype, "./data/chiptype.rds")
```

### Normalization of CEL Files

The following processes the CEL files from each chip type separately using the
MAS5 normalization algorithm. The results will be written to 3 subdirectores
under `data` that are named after the chip type names. To reduce the memory
consumption of this step, the CEL files are normalized in batches of 200. The
normalization takes about 10 hours without parallelization.  To save time, this
process can be easily accelerated on a computer cluster.

```{r normalize_chips, eval=FALSE}
chiptype <- readRDS("./data/chiptype.rds")
chiptype_list <- split(names(chiptype), as.character(chiptype))
normalizeCel(chiptype_list, batchsize=200, rerun=FALSE) 
```

Next the results from each chip type are assembled in a data frame. After this
all three of these data frames are combined to a single one, here named `mas5df`.
```{r comb_chip_type_data, eval=FALSE}
chiptype_dir <- unique(chiptype)
combineResults(chiptype_dir, rerun=FALSE)
mas5df <- combineNormRes(chiptype_dir, norm_method="MAS5")
```

### Gene Level Data
After moving the `myAnnot.rds` file from above into the 'data' directory, the `probe2gene`
function is used to transforms probe set to gene level data. If genes are represented by
several probe sets then their mean intensities are used.

```{r prof2gene, eval=FALSE}
myAnnot <- readRDS("./data/myAnnot.rds") 
mas5df <- probe2gene(mas5df, myAnnot) 
saveRDS(mas5df,"./data/mas5df.rds")
```

### Average Intensities Across Replicates

The following averages the normalized gene expression values for different
concentrations, treatment times and replicates of compounds and cell types.
```{r rma2cmap_expr, eval=FALSE}
mas5df <- readRDS("./data/mas5df.rds") # dim: 12403 x 7056
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
cmap_drug_cell_expr <- meanExpr(mas5df, cmap_inst) # dim: 12403 X 3587
saveRDS(cmap_drug_cell_expr, "./data/cmap_drug_cell_expr.rds")
```

### Mean Intensities in HDF5

The normalized and averaged expression values are saved to an HDF5 file,
referred to as `cmap_expr` database.

```{r gen_cmap_expr, eval=FALSE}
cmap_drug_cell_expr <- readRDS("./data/cmap_drug_cell_expr.rds")
## match colnames to '(drug)__(cell)__(factor)' format
colnames(cmap_drug_cell_expr) <- gsub("(^.*)_(.*$)", "\\1__\\2__trt_cp", 
                                      colnames(cmap_drug_cell_expr)) 
matrix2h5(cmap_drug_cell_expr, "./data/cmap_expr.h5", overwrite=TRUE)
## Read in cmap_expr.h5 by chunks
cmap_expr_se <- readHDF5chunk("./data/cmap_expr.h5", colindex=1:5)
```

# Custom Signature Databases

Custom databases of GESs can be built with the `build_custom_db` function
provided by the `signatureSearch` package. For this the user provides custom
genome-wide gene expression data (e.g. for drug, disease or genetic
perturbations) in a `data.frame` or `matrix`. The gene expression data can be
most types of the pre-processed gene expression values described under section
1.4 of the `signatureSearch` vignette.


# Additional Datasets

The `signatureSearchData` package also contains several annotation datasets, such
as drug-target and mode of action (MOA) information of small molecules. They are 
required for `signatureSearch's` functional enrichment analysis (FEA) routines. 
Currently, most of these annotation data were downloaded from the following databases:

+ [DrugBank](https://www.drugbank.ca/) 
+ [CLUE](https://clue.io/) 
+ [STITCH](http://stitch.embl.de/)


<!-- 

# Supplemental Material

## `cmap` signature database

The MAS5 normalized CEL files from the "cmap_expr signature database" section
can be used for DE analysis with `limma` package to get the `logMA` matrix 
containing the LFC scores. The treatment v.s. control instances 
were defined in the `cmap_instances_02.txt`. The same as the `cmap_expr` database,
only one treatment condition is selected for a compound in a cell. 
So, the resulting `logMA` matrix has LFC scores of 1,281 compound treatments in 5 cells
(3,478 signatures in total). The latter was stored in an HDF5 file, which is 
referred to as the `cmap` database. Note, The number of compound treatments in `cmap`
database is slightly different from that of the `cmap_expr` database. The reason
is that some of the compound treatment is discarded if the number of control and treatment
samples are less than 3 during the DE analysis.

### Load from `AnnotationHub`
The preprocessed `cmap` database can be loaded through the `AnnotationHub` interface as follows.
```{r cmap, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("signatureSearchData", "cmap"))
cmap_expr_path <- ah[["AH69090"]]
```

In summary, the loaded data instance includes LFC scores of 12,403
genes of 1,281 compound treatments across a total of 5 cell lines. This data set 
can be used by all GESS methods provided by the `signatureSearch` package.

### Generate from the original resources
The following steps explain how to generate the above data set from the MAS5 
normalized expression matrix of CEL files (`mas5df`) generated at the
"cmap_expr signature database" section. Use the same working directory as 
`cmap_expr` signature database.

#### Generate list of CEL files defining treatment vs. control comparisons

The `sampleList` function extracts the sample comparisons (contrasts) from the
CMAP annotation table and stores them as a list.
```{r cel_file_list, eval=FALSE}
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
comp_list <- sampleList(cmap_inst, myby="CMP_CELL")
```

#### DEG analysis with `limma`
The analysis of differentially expressed genes (DEGs) is performed with the `limma` package.
```{r deg_limma, eval=FALSE}
mas5df <- readRDS("./data/mas5df.rds")
degList <- runLimma(df=log2(mas5df), comp_list, fdr=0.10, foldchange=1, verbose=TRUE)
saveRDS(degList, "./results/degList.rds") # saves entire degList
```

#### Save the `logMA` into an HDF5 file
The `logMA` contains the LFC scores of compound treatments in cells. It is saved
to an HDF5 file, which is the `cmap` database.
```{r se, eval=FALSE}
degList <- readRDS("./results/degList.rds")
logMA <- degList$logFC
## match colnames of logMA to '(drug)__(cell)__(factor)' format
colnames(logMA) <- gsub("(^.*)_(.*$)", "\\1__\\2__trt_cp", colnames(logMA)) 
matrix2h5(logMA, "./data/cmap.h5", overwrite=TRUE) # 12403 X 3478
## Read in cmap.h5 by chunks
cmap_se <- readHDF5chunk("./data/cmap.h5", colindex=1:5)
```

-->

<font color="blue">[ThG-Comment: I have commented out the Supplement section
with the DEG analysis. Do we really need this section for anything in the
signatureSearch package, or the paper?]</font>

# Session Info

```{r sessionInfo}
sessionInfo()
```

# Funding

This project is funded by NIH grants
[U19AG02312](https://www.longevityconsortium.org/) and
[U24AG051129](https://www.longevitygenomics.org/) awarded by the National
Institute on Aging (NIA). Subcomponents of the environment are based on methods
developed by projects funded by NSF awards ABI-1661152 and PGRP-1810468. The
High-Performance Computing (HPC) resources used for optimizing and applying the the code of
this project were funded by NIH and NSF grants 1S10OD016290-01A1 and MRI-1429826,
respectively. 

# References
