---
title: _signatureSearchData_ - Reference Data for Large-Scale Gene Expression Signature Searching
author: "Authors: Yuzhu Duan and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 4
    fig_caption: yes
    code_folding: show
    number_sections: true
  pdf_document:
    toc: true
fontsize: 15pt
bibliography: bibtex.bib
vignette: >
  %\VignetteIndexEntry{signatureSearchData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
    library(signatureSearchData)
})
<<<<<<< HEAD
=======
# knitr::opts_knit$set(root.dir = "~/insync/project/longevityTools_eDRUG/")
>>>>>>> 21f15724563265e913ad353fe7fb7be317e6ce50
```

# Introduction

The `signatureSearchData` package provides access to the reference data used by
the associated `signatureSearch` software package. The latter allows to search 
with a query gene expression signature a database of treatment signatures to 
identify cellular states with similar expression responses (connections). This 
way one can identify drugs or gene knockouts that induce expression phenotypes 
similar to a sample of interest. The resulting associations may lead to novel functional insights how
perturbagens of interest interact with biological systems. Currently, `signatureSearchData` includes
gene expression data from the CMap (Connectivity Map) and LINCS (Library of
Network-Based Cellular Signatures) projects that are largely based on drug and
genetic perturbation experiments performed on variable numbers of human cell
lines [@Lamb2006-du; @Subramanian2017-fu]. In `signatureSearchData` these data sets 
are preprocessed as required for the gene expression signature search (GESS)
algorithms implemented and used by `signatureSearch`. The preprocessed data
types include but are not limited to normalized gene expression values 
(_e.g._ intensity values), log fold changes (LFC) and Z-scores, P or FDR values 
for DEG calls, rankings based on selected preprocessing routines or sets of 
top up/down-regulated DEGs. 

The CMap data were downloaded from the [CMap project
site](https://portals.broadinstitute.org/cmap/) (Version CMAP2). The latter is
a collection of over 7,000 gene expression profiles (signatures) obtained
from perturbation experiments with 1,309 drug-like small molecules on eight
human cancer cell lines. The Affymetrix Gene Chip technology was used to
generate the CMAP2 data set. 

In 2017, the LINCS Consortium generated a similar but much larger data set where
the total number of gene expression signatures was scaled up to over one
million. This was achieved by switching to a much more cost effective gene
expression profiling technology called L1000 assay [@Peck2006-rf;
@Edgar2002-di]. The current set of perturbations covered by the LINCS data set
includes 19,811 drug-like small molecules applied at varialbe concentrations
and treatment times to 70 human non-cancer (normal) and cancer cell lines.
Additionally, it includes several thousand genetic perturbagens composed of
gene knockdown and over-expression experiments. 

The data structures and search algorithms used by `signatureSearch` and
`signatureSearchData` are designed to work with most genome-wide expression
data including hybridization-based methods, such as Affymetrix or L1000, as
well as sequencing-based methods, such as RNA-Seq. Currently,
`signatureSearchData` does not include RNA-Seq data mainly due to the lack
of large-scale perturbation studies (_e.g._ drug-based) available in the public
domain that are based on RNA-Seq. This situation may change in the near future
once the cost of RNA-Seq becomes more cost effective for this purpose. 


# Install and load `signatureSearchData`

`signatureSearch` is a R/Bioconductor package and can be installed using 
`BiocManager::install()`.
```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("signatureSearchData", version = "3.9")
```

After the package is installed, it can be loaded in an R session as follows.
```{r load, eval=FALSE}
library(signatureSearchData)
```

# `lincs` signature database
<font color="blue">[Comment: start here with a brief and general description how 
the LINCS data is provided, e.g. the different levels of preprocessing available
for download. The next paragraph addresses this. So you may want to moved it here to 
address this. </font>
The `lincs` database represents moderated z-scores from DE analysis of 12,328 
genes from 8,140 compound treatments in 30 cells. Since one compound has treatments
in multiple cells, the database contains 45,956 signatures (drug treatments in cells)
in total. It can be used for all the GESS methods in the `signatureSearch` package.
<font color="blue">[Comment: neither I or others will be able to understand this paragraph.
You need to be more precise so that the numbers you mention here make sense.]</font>

## Load from `AnnotationHub`
The preprocessed `lincs` database can be loaded through the `AnnotationHub` interface

```{r lincs, eval=FALSE}
library(AnnotationHub); library(rhdf5)
ah <- AnnotationHub()
qr <- query(ah, c("signatureSearchData", "lincs"))
lincs_db_path <- ah[['AH69092']]
h5ls(lincs_db_path)
```

## Generate from the original resources

The LINCS data are available for download from GEO at 5 different pre-processing levels. 
Level 1 data are the raw mean fluorescent intensity (MFI) values that come 
directly from the Luminex scanner. Level 2 data are the expression intensities
of 978 landmark genes directly measured by the L1000 assay. They have been normalized and 
used to impute the expression of an additional 11,350 genes, forming Level 3 data. 
A robust z-scoring procedure was used to generate differential expression values from 
the normalized profiles (Level 4). Finally, a moderated z-scoring procedure was 
applied to the replicated samples of each experiment (generally 3 replicates) to compute a 
weighted average signature (Level 5). For a more detailed description of the preprocessing methods used by
the LINCS, readers want to refer to the [LINCS user guide](https://docs.google.com/document/d/1q2gciWRhVCAAnlvF2iRLuJ7whrGP6QjpsCMq1yWz7dU/edit#).

The steps in subsections <font color="blue">??? to ???</font> download LINCS' Level 5 data in `gctx` format, process and save 
them to an HDF5 file (here `lincs` database).  

### Download the original LINCS Level5 datasets at GEO 

Download the following files from GEO entry [GSE92742](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742):

+ GSE92742_Broad_LINCS_Level5_COMPZ.MODZ_n473647x12328.gctx.gz
+ GSE92742_Broad_LINCS_gene_info.txt.gz
+ GSE92742_Broad_LINCS_sig_info.txt.gz

Unzip the downloaded files to the `data` directory under the present working 
directory of the user’s R session. <font color="blue">User will not understand the purpose of the
data directory and its specific name. You need to explain this.</font>

<strike> The LINCS Level5 signatures represent moderated z-scores of compound treatments
at different concentration and time in different cells from differential expression
analysis, the replicates have been weighted averaged.</strike> <font color="blue">Duplicates
above. Stay things once and clearly. Don't keep duplicating since it will be confusing. If
you want to refernce then give things a short name! </font> 

### Filter signatures 

LINCS level5 dataset contains signatures from compound treatment in different 
cells at multiple concentration and duration as well as genetic perturbations. 
Since we only focus on signatures of compound/drug treatment in different cells, 
the concentration and duration information was ignored by only selecting those signatures at specific concentration (10 um) and duration (24 h) since most drugs have this dosage and time point. 
```{r filter_meta42, eval=FALSE}
## ignore warning message
meta42 <- readr::read_tsv("./data/GSE92742_Broad_LINCS_sig_info.txt") 
dose <- meta42$pert_idose[7]
## filter rows by pert_type as compound, 10 uM concentration, and 24 h treatment time
meta42_filter <- sig_filter(meta42, pert_type="trt_cp", dose=dose, time="24 h") # 45956 X 14
```

### Save z-score matrix from gctx file to HDF5 file 

Read in the large z-score matrix of treatments after filter from the gctx file by chunks
and write the matrix to an HDF5 file, which is the `lincs` database.
Since the size of the matrix would be tens of gigabytes, 
it will consume all the memories of a laptop if reading the 
full matrix at once. Reading the matrix by chunks and append to an HDF5 file will 
be memory efficient. 

We also provide the utility to read in subset of a large matrix in HDF5 file as 
`SummarizedExperiment` object by chunks/indexing.
```{r extract_modz, eval=FALSE}
gctx="./data/GSE92742_Broad_LINCS_Level5_COMPZ.MODZ_n473647x12328.gctx"
gctx2h5(gctx, cid=meta42_filter$sig_id, new_cid=meta42_filter$pert_cell_factor,
        h5file="./data/lincs.h5", chunksize=5000, overwrite=TRUE)
se <- readHDF5chunk(h5file="./data/lincs.h5", colindex=1:5000)
```

# `lincs_expr` signature database

The `lincs_expr` database contains mean gene expression values 
from 5,925 compound treatment samples in 30 cells (38,824 signatures in total). 

## Load from `AnnotationHub`
The preprocessed `lincs_expr` database can be loaded through the `AnnotationHub` interface
```{r lincs_expr, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
qr <- query(ah, c("signatureSearchData", "lincs"))
lincs_expr_path <- ah[['AH69093']]
```

## Generate from the original resources
The original LINCS Level3 expression database was downloaded at [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742).
It contains normalized expression values of 12328 genes from 1.3 million
treatment and control samples (compound treatment in different cells at various
concentration and duration, including replicates as well as genetic perturbations). 
The downloaded `gctx` file was processed and saved as an HDF5 file.

### Download the original LINCS Level3 datasets at GEO 

Download the following files from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742):

+ GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx.gz
+ GSE92742_Broad_LINCS_gene_info.txt.gz
+ GSE92742_Broad_LINCS_inst_info.txt.gz

Unzip the downloaded files to the `data` directory under the present working 
directory of the user’s R session.

### Filter signatures 

The same as the LINCS level5 data, filter the level3 signatures at specific 
concentration (10 um) and treatment time (24 h) 
```{r filter_expr, eval=FALSE}
inst42 <- readr::read_tsv("./data/GSE92742_Broad_LINCS_inst_info.txt") 
inst42_filter <- inst_filter(inst42, pert_type="trt_cp", dose=10, dose_unit="um", 
                             time=24, time_unit="h") # 169,795 X 13
```

### Calculate mean expression value and write the expression matrix to HDF5 file

Calculate mean expression values of replicated samples that are treated by the same
compound in the same cell type with selected concentration and treatment time.
Then write the mean expression matrix to an HDF5 file, which is the `lincs_expr`
database
```{r extract_expr, eval=FALSE}
# It takes some time
meanExpr2h5(gctx="./data/GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx",
            inst=inst42_filter, h5file="./data/lincs_expr.h5") # 12328 X 38824
```

# `cmap_rank` signature database
The `cmap_rank` database contains rank profiles of 12,403 genes from 1,281 compound 
treatments in 4 cells (3,478 signatures in total). It can be used to calculate
the connectivity scores to the query gene sets by `gess_cmap` method in the
`signatureSearch` package.

## Load from `AnnotationHub`
The preprocessed `cmap_rank` database can be loaded through the `AnnotationHub` interface
```{r cmap_rank, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
qr <- query(ah, c("signatureSearchData", "cmap"))
cmap_rank_path <- ah[['']]
```

## Generate from the original resources

The CMap group has pre-processed a `rankMatrix` storing the ranks of probe sets 
for each instance (a treatment and control pair and the probe sets were ordered 
by their extent of differential expression between this treatment and control pair).

The following steps will download, process the `rankMatrix` and store it in an
HDF5 file (`cmap_rank` database).

### Download the `rankMatrix` from [CMap 02](https://portals.broadinstitute.org/cmap)

Download the [rankMatrix.txt.zip](ftp://ftp.broad.mit.edu/pub/cmap/rankMatrix.txt.zip)
and unzip it to the `data` directory under the present working directory of the user’s R session.

### Filter instances

In order to get gene rank profiles from compound treatments in cells, the CMap 
instances were filtered by selecting only one instance of a compound treatment in a cell.
```{r filter_rankm, eval=FALSE}
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
inst_id <- cmap_inst$instance_id[!duplicated(paste(cmap_inst$cmap_name, cmap_inst$cell2, sep="_"))]
rankM <- read.delim("./data/rankMatrix.txt", check.names=FALSE, row.names=1)
rankM_sub <- rankM[,as.character(inst_id)]
colnames(rankM_sub) <- unique(paste(cmap_inst$cmap_name, cmap_inst$cell2, "trt_cp", sep="__"))
```

###  Transform probe set to gene level data
The probe sets were then transformed to gene levels by taking mean value of ranks of
probes representing the same gene.

#### Obtain annotation information
The following generates annotation information for the Affymetirx probe set
identifiers. Note, the three different Affymetrix chip types used by CMap
share most probe set ids (>95%), meaning it is possible to use the same 
annotation package. Create a `results` directory under your current working 
directory of R session if it does not exist.
```{r affyid_annot, eval=FALSE, message=FALSE}
library(hgu133a.db)
myAnnot <- data.frame(ACCNUM=sapply(contents(hgu133aACCNUM), paste, collapse=", "), 
                      SYMBOL=sapply(contents(hgu133aSYMBOL), paste, collapse=", "), 
                      UNIGENE=sapply(contents(hgu133aUNIGENE), paste, collapse=", "), 
                      ENTREZID=sapply(contents(hgu133aENTREZID), paste, collapse=", "), 
                      ENSEMBL=sapply(contents(hgu133aENSEMBL), paste, collapse=", "), 
                      DESC=sapply(contents(hgu133aGENENAME), paste, collapse=", "))
saveRDS(myAnnot, "./results/myAnnot.rds")
```

#### Transform probe sets to genes
```{r mr_prob, eval=FALSE}
rankM_sub_gene <- probe2gene(rankM_sub, myAnnot) # It takes some time
```

### Write the processed rank matrix to an HDF5 file

Write the the processed CMap rank matrix to an HDF5 file, which is the `cmap_rank`
database
```{r cmap_rank2h5, eval=FALSE}
matrix2h5(rankM_sub_gene, "./data/cmap_rank.h5", overwrite=TRUE) # 12403 X 3587
rhdf5::h5ls("./data/cmap_rank.h5")
## Read in cmap_rank.h5 as SummarizedExperiment object by chunks
cmap_rank_se <- readHDF5chunk("./data/cmap_rank.h5", colindex=1:5)
```

# `cmap_expr` signature database
For correlation based GESS methods (`gess_cor`) in the `signatureSearch` software 
package, the signature database containing mean gene expression intensity values of drug 
treatment samples could also be used to search for similarity when the query
signature are also gene expression values. 

The `cmap_expr` database represents mean expression values of 1,309 drug
treatment samples in 4 cells (3,587 signatures in total).

## Load from `AnnotationHub`
The preprocessed `cmap_expr` database is stored in an HDF5 file and can be 
loaded through the `AnnotationHub` interface
```{r cmap_expr, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
qr <- query(ah, c("signatureSearchData", "cmap_expr"))
cmap_expr_path <- ah[["AH69091"]]
```

## Generate from the original resources
The source data was downloaded from the 
[Connectivity Map](https://portals.broadinstitute.org/cmap/) project, 
which are the CEL files generated by Affymetrix chips. 
The downloaded CEL files were normalized by `MAS5` method. 
Then, the expression values of drug treatment samples at different 
concentration, duration were averaged and used as expression values of that drug
treatment in a cell type. The expression matrix was saved in an HDF5 file, which
is the `cmap_expr` database. The following steps were used to generate the 
`cmap_expr` database from CEL files.

### Create expected directory structure
The following creates the expected directory structure. Input data will be 
stored in the data directory and results will be written to the results 
directory. All paths are given relative to the present working directory of 
the user’s R session.
```{r work_envir, eval=FALSE}
dir.create("data"); dir.create("data/CEL"); dir.create("results") 
```

### Download data from Connectivity Map project site
The drug-related expression data are downloaded from the CMAP web site
[here](http://www.broadinstitute.org/cmap). The `getCmapCEL` function
downloads the 7,056 CEL files. The functions will write the
downloaded files to the `data/CEL` directories within the present 
working directory of the user's R session. Since some of the raw data sets 
are large, the functions will only rerun the download if the argument `rerun` 
is assigned `TRUE`. If the raw data are not needed then users can skip this 
time consuming download step and work with the preprocessed data 
obtained in the next section.

```{r download_cmap, eval=FALSE}
getCmapCEL(rerun=FALSE) # Download cmap CEL files. this will take hours
```

The experimental design of the CMAP project is defined in the file
`cmap_instances_02.xls`.  Note, this file required some cleaning in LibreOffice
(Excel would work for this too). After this it was saved as tab delimited txt
file named `cmap_instances_02.txt`. It is available from this package.
```{r cmap_inst2, eval=TRUE}
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
```

The panel of cell lines used by CMAP includes 
[MCF7](http://www.broadinstitute.org/cmap/help_topics_linkified.jsp#mcf7), 
[ssMCF7](http://www.broadinstitute.org/cmap/help_topics_linkified.jsp#mcf7), 
[HL60](http://www.broadinstitute.org/cmap/help_topics_linkified.jsp#hl60), 
[PC3](http://www.broadinstitute.org/cmap/help_topics_linkified.jsp#pc3) and 
[SKMEL5](http://www.broadinstitute.org/cmap/help_topics_linkified.jsp#skmel5). 

### Determine chip type from CEL files 
The CMAP data set is based on three different Affymetrix chip types (HG-U133A,
HT_HG-U133A and U133AAofAv2). The following extracts the chip type information
from the CEL files and stores the result in an `rds` file with the path 
`./data/chiptype.rds`.

```{r get_cel_type, eval=FALSE}
library(affxparser)
celfiles <- list.files("./data/CEL", pattern=".CEL$")
chiptype <- sapply(celfiles, function(x) affxparser::readCelHeader(paste0("data/CEL/", x))$chiptype)
saveRDS(chiptype, "./data/chiptype.rds")
```

### Normalization of CEL files

The following processes the CEL files from each chip type separately using the
MAS5 normalization algorithm. The results will be written to 3 subdirectores
under `data` that are named after the chip type names. Since thousands of CEL 
files could not be read into memory at once, they are normalized in batch
by processing 200 files at once. The normalization process takes about 10 hours, 
to save time, it can be easily paralleled to run multiple jobs at the same time
on a computer cluster with a scheduler (e.g. Slurm).

The number of CEL files from each chip type are: 807 CEL files from HG-U133A, 
6029 CEL files from HT_HG-U133A, and 220 CEL files from U133AAofAv2. 
Note, these numbers are slightly different than those reported in the 
`cmap_instances_02.txt` file. 

```{r normalize_chips, eval=FALSE}
chiptype <- readRDS("./data/chiptype.rds")
chiptype_list <- split(names(chiptype), as.character(chiptype))
normalizeCel(chiptype_list, batchsize=200, rerun=FALSE) # it takes about 10 hours
```

Combine results from same chip type in single data frame, then combine the data
frame from different chip types into `mas5df`
```{r comb_chip_type_data, eval=FALSE}
chiptype_dir <- unique(chiptype)
combineResults(chiptype_dir, rerun=FALSE)
mas5df <- combineNormRes(chiptype_dir, norm_method="MAS5")
```

### Transform probe set to gene level data
Copy the `myAnnot.rds` generated from the above "cmap_rank signature database" 
session to the `result` directory under your current R session
```{r prof2gene, eval=FALSE}
myAnnot <- readRDS("./results/myAnnot.rds") 
mas5df <- probe2gene(mas5df, myAnnot) # It takes about 15 minutes
saveRDS(mas5df,"./data/mas5df.rds")
```

### Take mean expression values of drug treatment samples

Take mean expression values of multiple drug treatment samples at different 
concentration and duration in the same cell as expression value of that drug
treatment in the cell.
```{r rma2cmap_expr, eval=FALSE}
mas5df <- readRDS("./data/mas5df.rds") # matrix: 12403 7056
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
cmap_drug_cell_expr <- meanExpr(mas5df, cmap_inst) # 12403 X 3587
saveRDS(cmap_drug_cell_expr, "./data/cmap_drug_cell_expr.rds")
```

### Store the expression matrix to an HDF5 file
```{r gen_cmap_expr, eval=FALSE}
cmap_drug_cell_expr <- readRDS("./data/cmap_drug_cell_expr.rds")
## match colnames to '(drug)__(cell)__(factor)' format
colnames(cmap_drug_cell_expr) <- gsub("(^.*)_(.*$)", "\\1__\\2__trt_cp", 
                                      colnames(cmap_drug_cell_expr)) 
matrix2h5(cmap_drug_cell_expr, "./data/cmap_expr.h5", overwrite=TRUE)
## Read in cmap_expr.h5 by chunks
cmap_expr_se <- readHDF5chunk("./data/cmap_expr.h5", colindex=1:5)
```

# Discussion
The `lincs`, `lincs_expr`, `cmap_rank`, and `cmap_expr` signature databases we
generated above contain genome-wide expression profiles of drug treatments in
cells. The profile could be z-scores from differential expression analysis, gene
ranks, expression intensity values from microarray, log transformed read counts 
from RNA-seq *etc*. Users can also build their custom database via 
`build_custom_db` function in the `signatureSearch` software package. 

The signature database is technology independent, the profiles could be from
Affymetrix chips, L1000 assay, RNA-seq or other gene expression profiling 
technology. The treatments could be any of bioactive small molecules, genetic 
perturbations, disease state *etc*.

For the CMap dataset, the MAS5 normalized expression data could also be used for
DE analysis, the resulting scores (e.g. log2 fold change) can be used to build
the `cmap` signature database used for all the GESS methods in the 
`signatureSearch` package. The details is under the "Supplement" section.

# Other datasets
This annotation package also contains other intermediate datasets used for
functionality in the `signatureSearch` software package, 
such as getting targets of query drugs in [DrugBank](https://www.drugbank.ca/), 
[CLUE](https://clue.io/) and [STITCH](http://stitch.embl.de/) databases, 
conducting GO enrichment analysis or computing WTCS p-values or tau scores of the
`gess_lincs` method in the `signatureSearch` package. 

# Supplement

## `cmap` signature database
The `cmap` database represents the $log_2$ fold change (log2FC) of 12,403 genes 
from 1,281 compound treatments in 4 cells (3,478 signatures in total).

### Load from `AnnotationHub`
The log2FC matrix is stored in an HDF5 file and can be loaded through the 
`AnnotationHub` interface
```{r cmap, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
qr <- query(ah, c("signatureSearchData", "cmap"))
cmap_expr_path <- ah[["AH69090"]]
```

### Generate from the original resources

The MAS5 normalized expression matrix ('mas5df') from the 
"cmap_expr signature database" session was used for DE analysis with `limma` 
package to get the `logMA` matrix containing the log2 fold change scores. 
The `logMA` matrix was stored in an HDF5 file, which is the `cmap` database.
Use the same working directory as `cmap_expr` signature database.

#### Generate list of CEL names defining treatment vs. control comparisons

The `sampleList` function extracts the sample comparisons (contrasts) from the
CMAP annotation table and stores them as a list.
```{r cel_file_list, eval=FALSE}
path <- system.file("extdata", "cmap_instances_02.txt", package="signatureSearchData")
cmap_inst <- read.delim(path, check.names=FALSE) 
comp_list <- sampleList(cmap_inst, myby="CMP_CELL")
```

#### DEG analysis with `limma`

The analysis of differentially expressed genes (DEGs) is performed with the `limma` package.
The log2 fold change (log2FC) scores of drug treatments in cells are used as 
gene expression profiles. 

```{r deg_limma, eval=FALSE}
mas5df <- readRDS("./data/mas5df.rds")
degList <- runLimma(df=log2(mas5df), comp_list, fdr=0.10, foldchange=1, verbose=TRUE)
saveRDS(degList, "./results/degList.rds") # saves entire degList
```

#### Store the `logMA` into an HDF5 file
The generated HDF5 file is the `cmap` database storing the log2FC scores of 
compound treatments in cells.
```{r se, eval=FALSE}
degList <- readRDS("./results/degList.rds")
logMA <- degList$logFC
## match colnames of logMA to '(drug)__(cell)__(factor)' format
colnames(logMA) <- gsub("(^.*)_(.*$)", "\\1__\\2__trt_cp", colnames(logMA)) 
matrix2h5(logMA, "./data/cmap.h5", overwrite=TRUE) # 12403 X 3478
## Read in cmap.h5 by chunks
cmap_se <- readHDF5chunk("./data/cmap.h5", colindex=1:5)
```

# References
